Rectangulation polytopes

### 2024-03-12 draft/rectangulotopes.tex
formulas diagonal insertion

@@

\begin{figure}
	\centerline{\includegraphics[scale=.7]{weakRectangulation} \qquad \includegraphics[scale=.7]{weakRectangulationTrees}}
	\caption{A weak rectangulation (left) and its pair of separated twin trees (right). The corresponding cotwisted Baxter permutation is $[7,14,15,16,8,11,13,10,5,6,1,4,9,2,3,12]$ and its inverse is $[11, 14, 15, 12, 9, 10, 1, 5, 13, 8, 6, 16, 7, 2, 3, 4]$.}
\end{figure}

We call \defn{antidiagonal} the segment joining the top left corner to the bottom right corner.
A rectangulation~$R$ is \defn{diagonal} (or \defn{weak}) if it satisfies the following equivalent conditions:
\begin{enumerate}[(i)]
\item all rectangles of~$R$ meet the antidiagonal,
\item the source~$s(r)$ (resp.~target~$t(r)$) of each rectangle~$r$ of~$R$ is below (resp.~above) the antidiagonal,
\item the source tree~$S(R)$ (resp.~target tree~$T(R)$) remains below (resp.~above) the antidiagonal,
\item the source tree~$S(R)$ and the target tree~$T(R)$ do not overlap along a segment of~$R$,
\item $R$ avoids the patterns \horizontalPattern{} and \verticalPattern{},
\item the $2$-clumped permutation~$\projDown(R)$ avoids the patterns~$2413$ and~$3412$,
\item the co-$2$-clumped permutation~$\projDown(R)$ avoids the patterns~$2143$ and~$3142$,
\end{enumerate}

>>>>

%For a permutation~$\sigma$ of~$[n]$ and~$i \in [n]$, we define
%\begin{alignat*}{2}
%pv(i, \sigma) & \eqdef p(i) \cap v(i, \sigma), &
%pV(i, \sigma) & \eqdef p(i) \cap V(i, \sigma), \\
%Pv(i, \sigma) & \eqdef P(i) \cap v(i, \sigma) & \qquad\text{and}\qquad
%PV(i, \sigma) & \eqdef P(i) \cap V(i, \sigma).
%\end{alignat*}
%where~$p(i) \eqdef [i-1]$ while~$P(i) \eqdef [n] \ssm [i]$, and~$v(i, \sigma) \eqdef \sigma([i-1])$ while $V(i, \sigma) \eqdef \sigma([n] \ssm [i])$.

@@

\begin{figure}
	\centerline{\includegraphics[scale=.7]{weakRectangulation} \qquad \includegraphics[scale=.7]{weakRectangulationTrees}}
	\caption{A weak rectangulation (left) and its pair of separated twin trees (right). The corresponding twisted Baxter permutation is $[7, 5, 1, 14, 8, 6, 4, 2, 11, 10, 9, 3, 15, 16, 13, 12]$ and its inverse is $[3, 8, 12, 7, 2, 6, 1, 5, 11, 10, 9, 16, 15, 4, 13, 14]$.}
\end{figure}

We call \defn{diagonal} the segment joining the top left corner to the bottom right corner.
A rectangulation~$R$ is \defn{diagonal} (or \defn{weak}) if it satisfies the following equivalent conditions:
\begin{enumerate}[(i)]
\item all rectangles of~$R$ meet the diagonal,
\item the source~$s(r)$ (resp.~target~$t(r)$) of each rectangle~$r$ of~$R$ is below (resp.~above) the diagonal,
\item the source tree~$S(R)$ (resp.~target tree~$T(R)$) remains below (resp.~above) the diagonal,
\item the source tree~$S(R)$ and the target tree~$T(R)$ do not overlap along a segment of~$R$,
\item $R$ avoids the patterns \horizontalPattern{} and \verticalPattern{},
\item the $2$-clumped permutation~$\projDown(R)$ avoids the patterns~$2413$ and~$3412$,
\item the co-$2$-clumped permutation~$\projDown(R)$ avoids the patterns~$2143$ and~$3142$,
\end{enumerate}

\begin{proposition}
The weak rectangulation insertion of any permutation~$\sigma$ of~$[n]$ is given by the rectangles~$[w\ell(i, \sigma), wr(i, \sigma)-1] \times [n-wd(i, \sigma)+1, n-wu(i, \sigma)]$ for~$i \in [n]$ where
\begin{align*}
w\ell(i, \sigma) & \eqdef \max \; \{0\} \cup \set{j}{j < i \text{ and } \sigma^{-1}(j) < \sigma^{-1}(i)} \\
wr(i, \sigma) & \eqdef \min \; \{n+1\} \cup \set{j}{i < j \text{ and } \sigma^{-1}(i) < \sigma^{-1}(j)} \\
wd(i, \sigma) & \eqdef \min \; \{n+1\} \cup \set{j}{i < j \text{ and } \sigma^{-1}(j) < \sigma^{-1}(i)} \\
\text{and} \qquad
wu(i, \sigma) & \eqdef \max \; \{0\} \cup \set{j}{j < i \text{ and } \sigma^{-1}(i) < \sigma^{-1}(j)}.
\end{align*}
\end{proposition}
